### 2021-02-24 rad-icfp-2021/intro.tex
minor

In 2008, Pearlmutter and Siskind \cite{pearlmutter2008reverse} introduced AD in a functional setting. 
They introduced reverse-mode in their setting. 
Reverse-mode can compute in one pass the whole gradient of a function $\RR^n\to\RR$, and this is very fast.
By comparison, the easier forward-mode transformation would need $n$ passes. 
These functions $\RR^n\to\RR$ actually appear a lot in practice, for instance as loss functions in machine learning.
Reverse-mode is thus important as it computes a whole gradient very quickly for these functions. 
To be as efficient as the usual imperative version of reverse-mode, the transformation in \cite{pearlmutter2008reverse} introduced references.

@@

@@

\begin{example}[First-order term]
On simple first-order terms, the output code looks like SSA code \cite{}.
Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> Id(y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
% \begin{align*}
%     &\letin{w_1,w_1'}{\tPair{\var_1 * \var_2}{\inllambda \var[2]_1...\var[2]_4.Id(\var[2]_1+\var_2*\var[2]_4,\var[2]_2+\var_1*\var[2]_4,\var[2]_3)} }{
%     \\
%     &\letin{w_2,w_2'}{\tPair{w_1 * \var_1}{\inllambda \var[2]_1...\var[2]_5.w_1'(\var[2]_1+w_1 * \var[2]_5,\var[2]_2,\var[2]_3,\var[2]_4+ \var_1 * \var[2]_5)}}{\\ 
%     & w_2'(0,0,0,0,1)}}
% \end{align*}
where $Id$ is the identity function $\RR^3\to\RR^3$, $\inllambda$ is just a normal $\lambda$ tagged for inlining.
The part $(0,0,0,0,1)$ comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

After some general partial evaluation that will be detailed further in the paper, we obtain     
    	% \begin{align*} 
    	% 	&\letin{w_1}{\var_1 * \var_2}{\\
    	% 	&\letin{w_2}{w_1 * \var_1}{\\
    	% 	&\letin{\var[2]_1}{0}{\\
    	% 	&\letin{\var[2]_2}{0}{\\
    	% 	&\letin{\var[2]_3}{0}{\\
    	% 	&\letin{\var[2]_4}{0}{\\
    	% 	&\letin{\var[2]_5}{1}{\\
    	% 	& \letin{\var[2]_1'}{\var[2]_1+w_1 * \var[2]_5}{\\
    	% 	&\letin{\var[2]_4'}{\var[2]_4+ \var_1 * \var[2]_5}{\\
    	% 	&(\var[2]_1'+\var_2*\var[2]_4',\var[2]_2+\var_1*\var[2]_4',\var[2]_3)
    	% 	}}}}}}}}}
    	% \end{align*}
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
This is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.
Of course this can be further optimised to give
    	% \begin{align*}
    	% 	&\letin{w_1}{\var_1 * \var_2}{\\
    	% 	&\letin{w_2}{w_1 * \var_1}{\\
    	% 	&(w_1+\var_2*\var_1,\var_1*\var_1,0)
    	% 	}}
    	% \end{align*}
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

>>>>

In 2008, Pearlmutter and Siskind \cite{pearlmutter2008reverse} introduced Automatic Differentiation (AD) in a functional setting. 
They introduced reverse-mode in their setting. 
Reverse-mode can compute in one pass the whole gradient of a function $\RR^n\to\RR$, and this is very fast.
By comparison, the easier forward-mode transformation would need $n$ passes. 
These functions $\RR^n\to\RR$ actually appear a lot in practice, for instance as loss functions in machine learning.
Reverse-mode is thus important as it computes a whole gradient very quickly for these functions. 
To be as efficient as the usual imperative version of reverse-mode, the transformation in \cite{pearlmutter2008reverse} introduced references.

@@

\MH{forward reference to section 2 where reverse-mode is recalled. maybe same for SSA. }

@@

\begin{example}[First-order term]
On simple first-order terms, the output code looks like SSA code \cite{}.
Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part $(0,0,0,0,1)$ comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

After some general partial evaluation that will be detailed further in the paper, we obtain     
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
This is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.
Of course this can be further optimised to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}
